---
title: "Bonanza Hares Exploratory Report"
author: "Bri Baker"
date: "November 15, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(lubridate)
library(kableExtra)
library(effsize)
library(broom)
```


```{r import}
bonanza_hares <- read_csv(here("data" , "bonanza_hares.csv")) #load data
bonanza_juvenile <- bonanza_hares %>% # start with full dataset
  filter(age == "j") # filter juveniles
```

# Introduction

  The following exploration provides initial analysis of juvenile snowshoe hares, *Lepus americanus*, trapped in capture-recapture studies at the Bonanza Creek LTER in Interior Alaska. This keystone prey species is integral to the functioning of northern boreal forests, the primary ecosystem in Bonanza Creek. Previously, this data set was utilized to investigate the declining population of snowshoe hares in the area (Kielland et al., 2017). The focus of this exploration is to discern general relationships between juvenile hare size (determined by weight and hind foot length) and sex throughout the recorded time frame. 


# Data and Analysis

  The data utilized in this exploration are snowshoe hare trap data from the Bonanza Creek LTER taken from 1998 to 2012 (Kielland et al., 2017). Following are basic analyses of the juvenile snowshoe hare population. These analyses include explorations of population through time, weight comparisons by sex, and weight and hind foot comparisons. Methods utilized include Welch's two sample t-test, Cohen's d, and basic linear regression modeling in addition to basic graphical visualization.
  


# Exploratory Findings


### Annual Juvenile Trap Counts

```{r}
annual_juvenile <- bonanza_juvenile %>% # start with juvenile hares
  mutate(year_new = as.numeric(year(mdy(date)))) %>% # make year column as a number
  count(year_new) %>% # summarize individuals by years
  rename(count = n) # rename counts column

ggplot(annual_juvenile, aes(year_new, count)) + # make plot of year counts
  geom_line() + # use line graph
  scale_x_continuous("Year", breaks = seq(1998, 2012, 1)) + # format x axis
  theme_light() + # format theme
  labs(caption = "Figure 1. Annual count of juvenile showshoe hares at Bonanza LTER. Includes absoulte count \n of individuals per year." , # title
       y = "Individuals (count)") + # y label
  theme(panel.border = element_blank(), # remove panel border
        panel.grid.major = element_blank(), # remove major gridlines
        panel.grid.minor = element_blank(), # remove minor gridlines
        axis.line = element_line(colour = "black"), # make axis black
        plot.caption = element_text(hjust = 0, size = 11)) # move and size figure caption

juvenile_summary <- annual_juvenile %>% # start with juvenile
  summarize(mean_juv = mean(count), # summary table with mean
            median_juv = median(count), # median
            total_juv = sum(count), # sum
            max_trap = max(count), # max
            min_trap = min(count)) # min

```


  From 1998 to 2012, an average of `r juvenile_summary[1:1]` individuals were trapped annually with a maximum of `r juvenile_summary[1:4]` individuals and a minimum of `r juvenile_summary[1:5]` individuals which indicates that the measure of central tendency does not accurately represent the data. In part this is because over time the number of individuals trapped annually decreased (Figure 1). However, it is also important to note that these measures utilized absolute counts, so further analyses should standardize the counts by dividing by the total days of trapping in each year to ensure that the effort signal does not introduce variance that does not exist. 



### Juvenile Weights

```{r}
juvenile_weights <- bonanza_juvenile %>% # start with juvenile hares
  select(weight, sex, grid) %>% # select desired variables
  group_by(sex, grid) # group by sex and grid site

ggplot(juvenile_weights, aes(sex, weight)) + # plot juvenile weights
  geom_boxplot(aes(sex, weight)) + # boxplot
  geom_jitter(width = 0.25) + # jitter plot
  facet_wrap(~grid) + #split up by grid
  labs(caption = "Figure 2. Juvenile snowshoe hare weight (g) and sex comparison by site. Exloration includes \n individuals not identified by sex. Sites are seperated by trapping grid: bonbs (general trapping \n site), bonmat (traps set in mature forest), and bonrip (traps set in riparian flood zone). Points \n indicate individuals trapped.", 
       x = "Sex",
       y = "Weight (g)") +
  theme(panel.border = element_blank(), # remove panel border
        panel.grid.major = element_blank(), # remove major gridlines
        panel.grid.minor = element_blank(), # remove minor gridlines
        axis.line = element_line(colour = "black"), # make axis black
        plot.caption = element_text(hjust = 0, size = 11)) # move and size figure caption


```


  Hares trapped at the general site (bonbs) were larger on average than at the other two sites; however, there were more individuals and a larger spread trapped at the riparian site (bonrip). Across all trap sites, juvenile male snowshoe hare are larger on average than female juveniles. This difference is most defined in the mature forest trap site (bonmat) and least defined in the general trap site (bonbs) (Figure 2). 


### Juvenile Weight Comparison


```{r}
comparison_juveniles <- juvenile_weights %>% # start with juvenile weights
  filter(sex != "na") %>% # remove na
  group_by(sex) %>% # group by sex
  summarize(mean_weight_g = mean(weight, na.rm = TRUE), # find mean weight
            sd_weight_g = sd(weight, na.rm = TRUE), # find SD
            sample_size = n()) # count sample size
  
kable(comparison_juveniles, "html", # make a table in html
      caption = "Table 1. Summary statistics for weight comparison of juvenile snowshoe hare at Bonanza Creek LTER.",
      col.names = c("Sex", "Mean (g)", "SD (g)", "Sample Size")) %>% # rename columns
  kable_styling("striped") #use striped styling
```

```{r, include = FALSE}
## histogram and qq plot to check normality before running t-test
ggplot(juvenile_weights, aes(weight)) + # make ggplot of weight
  geom_histogram(binwidth = 55) + # use histogram
  facet_wrap(~sex) # split by sex

ggplot(juvenile_weights, aes(sample = weight)) + # make ggplot of weight
  geom_qq() + # use QQ
  facet_wrap(~sex) # split by sex

## Based on normal distributions indicated above, a t-test is appropriate
hares_f <- juvenile_weights %>% 
  filter(sex == "f") %>% 
  pull(weight) # make female vector

hares_m <- juvenile_weights %>% 
  filter(sex == "m") %>% 
  pull(weight) # make male vector

weight_ttest <- t.test(hares_f, hares_m) #ttest

weight_effect <- cohen.d(na.omit(hares_f), na.omit(hares_m)) # effect size


```

  The actual difference in means between the weights of juvenile snowshoe hares at the Bonanza Creek LTER is `r abs(round(comparison_juveniles[1,2] - comparison_juveniles[2,2], digits = 3))` g (calculated from Table 1). The p-value calculated using a `r weight_ttest$method` (p = `r round(weight_ttest$p.value, digits = 3)`) indicates a high likelihood that the two groups were drawn from samples with different means (i. e. there is a significant difference). However, with an estimated effect of `r round(weight_effect$estimate, digits = 3)`, the effect size calculated using `r weight_effect$method` is `r weight_effect$magnitude`.



### Juvenile Weight and Hind Foot Length

```{r}
ggplot(bonanza_juvenile, aes(weight, hindft)) + # plot weight and hindft
  geom_point() + # scatterplot
  stat_smooth(method = "lm", col = "black") +
  theme_classic() +
  labs(x = "Hind Foot Lenght (mm)",
       y = "Weight (g)",
       caption = "Figure 3. Relationship between weight (g) and hind foot lenght (mm) of juvenile snowshoe hares \n with results of linear regression model.") +
  theme(plot.caption = element_text(hjust = 0, size = 11)) # move and size figure caption
```

```{r}
lm_j_ft_weight <- lm(hindft ~ weight, data = bonanza_juvenile) # linear model

ft_weight_lm_tidy <- broom::tidy(lm_j_ft_weight) # make table

bonanza_int <- ft_weight_lm_tidy$estimate[1] # intercept

bonanza_coef <- ft_weight_lm_tidy$estimate[2] # coefficient

bonanza_lm_out <- broom::glance(lm_j_ft_weight) # stats glance

```


  Simple linear regression was used to explore the relationship between juvenile snowshoe hare foot hind foot length (mm) and body mass (g) across all three sampling sites, and including both male and female hares. A significant regression model was found ($\beta$ = `r round(bonanza_coef,3)`, F(`r bonanza_lm_out$df`,`r bonanza_lm_out$df.residual`) = `r round(bonanza_lm_out$statistic,1)`, p < 0.001) with an R^2^ of `r round(bonanza_lm_out$r.squared,3)`.

  The low R^2^ indicates that this linear model may not be the best fit for the patterns in these data. The data do not fit all of the main assumptions for a linear regression. Firstly, although there is a positive relationship in the data and in the model, it is difficult to say if the data fit the linearity assumption. Additionally, they are heteroscedastic and therefore do not fit the homoscedasticity assumption. 



# Summary

* In recent years the number of juvenile snowshoe hares captured annually has decreased. 
* Among juvenile hares, males tend to weigh more with average weights varying across trap sites. 
* The difference in weight between males and females is statistically significant, but the effect size of the phenomenon is small.
* There is a positive relationship between juvenile hare weight and hind foot length, but a linear model does not effectively explain the variance.



# Citations


Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. *Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative.* https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14

